---
title: Using variants unfiltered for MAF for PCA and heritability analysis
author: Martha Aquino
date: '2018-11-26'
slug: using-variants-unfiltered-for-maf-for-pca-and-heritability-analysis
categories:
  - Coding
  - Experiments
tags:
  - Data Cleaning
  - Genomic Data
  - Heritability
  - PCA
---

#Introduction 


#Methods and Results 

####1. Run PCA with 20 PCs on unfiltered MAF data set
```{bash, eval = FALSE}
plink1.9 --bfile merged_nies/merged_nies_geno_210818 --pca 20 var-wts --out plink_output/nies_geno_pca261118
```

Note:

- the merged_nies_geno_210818 contains the data set after HWE and missingness filters (contains 3.9 million variants).

- 20 PCs included. From the previous entry, it seemed that including 20PCs produced consistently higher heritability estimates than including 1 - 5 PC. 

####2. Generate PCA plots 

Load .eigenvec file 
```{r}
nies_geno_eigenvec261118 <- read.table('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/plink_output/nies_geno_pca261118.eigenvec', header = F)

```

Plot PC1 Vs PC2
```{r}
plot(nies_geno_eigenvec261118$V3, nies_geno_eigenvec261118$V4, xlab = "PC1", ylab = "PC2", main = "PC1 vs PC2 eigenvectors") 
```

####3. Make GRM 
```{bash, eval = FALSE}
./gcta64 --bfile merged_nies/merged_nies_geno_210818 --make-grm --out gcta_output/merged_nies_grm31218
```

####4. Perform PCA with 20 PCs
```{bash, eval = FALSE}
./gcta64 --bfile merged_nies/merged_nies_geno_210818 --pca 20 --out gcta_output/merged_nies_pca31218
```

```{r}
merged_nies_eigenvec31218 <- read.table('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/gcta_output/merged_nies_pca31218.eigenvec', header = F)

write.csv(merged_nies_eigenvec31218, 'C:/Users/Martha/Documents/Honours/Project/honours.project/Data/gcta_output/merged_nies_eigenvec31218.csv')
```


####5. Estimate heritability of individual traits 
```{bash, eval = FALSE}
for i in {1..27}
do ./gcta64 --reml --grm gcta_output/merged_nies_grm31218 --pheno nies_pheno_020918.txt --mpheno $i --reml-pred-rand --covar nies_sex_covar.txt --qcovar nies_cont_covar31218.txt --out gcta_output/heritability_031218/$i.est
done
```

####6. Estimate heritability of principal components
```{bash, eval = FALSE}
for i in {1..9}
do ./gcta64 --reml --grm gcta_output/merged_nies_grm31218 --pheno pheno_pca_coord.txt --mpheno $i --reml-pred-rand --covar nies_sex_covar.txt --qcovar nies_cont_covar31218.txt --out gcta_output/heritability_031218/$i.pc.est
done
```

####7. View heritability estimates 
```{r}
heritability_031218 <- read.csv('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/gcta_output/heritability_031218/new_heritability31218.csv', header = F)

heritability_031218
```

These heritability estimates are again different from the previous results. The differences are most likely due to the inclusion and exclusion of rare and low frequency variants. 

American Journal of Human Genetics - for publishing paper
PLOS Genetics 