---
title: Performing PCA on genomic data
author: Martha Aquino
date: '2018-08-08'
slug: performing-pca-on-genomic-data
categories:
  - Coding
  - Experiments
tags:
  - Genomic Data
  - Linux
  - PCA
  - PLINK
  - R
---

#Introduction
The final step to the second part of aim 1 is to perform a PCA on the genomic data to investigate the population structure. This is important because it will allow me to determine if there are any underlying genetic structures in the genomic data and I can account for relatedness among individuals for subsequent analyses. 

#Methods and Results 

###1. Run PCA 
```{bash, eval = F}
plink1.9 --bfile merged_nies --pca 180 var-wts --out plink_output/nies_final_pca
```
The count had to be increased (default is 20) because there are over 100 components with an eigenvalue >1. 

Including the var-wts modifier will generate a file for variant weights. 

###2. Load eigenvalue results 
```{r}
nies_final_pca_eigenval <- read.table('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/plink_output/nies_final_pca.eigenval', header = F)

head(nies_final_pca_eigenval)
```

###3. Produce screeplot
```{r}
barplot(nies_final_pca_eigenval$V1, 
        names.arg = 1:nrow(nies_final_pca_eigenval), 
        main = "NIES PCA Eigenvalue",
        xlab = "Principal Components",
        ylab = "Eigenvalue",
        col ="lightskyblue2")
lines(x = 1:nrow(nies_final_pca_eigenval), nies_final_pca_eigenval$V1, 
      type = "b", pch = 19, col = "red")
```

There are 124 principal components that have an eigenvalue >1. 



