---
title: Extracting common variants from both data sets
author: Martha Aquino
date: '2018-07-03'
slug: extracting-common-variants-from-both-data-sets
categories:
  - Coding
  - Experiments
tags:
  - Data Cleaning
  - Genomic Data
  - Linux
  - PLINK
  - PCA
  - R
---

#Introduction
The merged data set did not produce the expected output. I expected that the merging would only retain data from common variants between both genomic data sets. To resolve this, I need to compare the two data sets and extract the list of common variant IDs and use this to extract the relevant data. 

#Methods
1.Load .bim files for both data sets 
```{r, cache=TRUE}
wgs_bim_data <- read.table('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/NI_wgs_merged_snps.bim', header = F)

head(wgs_bim_data)
```

```{r, cache=TRUE}
snp_bim_data <- read.table('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/NImerged.impute2.chrAllX.2014-Oct-02.bim', header = F)

head(snp_bim_data)
```

2.Merge files to obtain list of common variants 
```{r, cache=TRUE}
merged_bim_data <- merge(wgs_bim_data, snp_bim_data, by = 'V2')

head(merged_bim_data)
```

3.Extract list of common variant IDs
```{r, cache=TRUE}
merged_rsID <- merged_bim_data$V2

head(merged_rsID)
```

4.Export list to a .txt file 
```{r, eval=FALSE}
write.table(merged_rsID, file = 'C:/Users/Martha/Documents/Honours/Project/honours.project/Data/merged_rsID.txt', sep = "\t", row.names = F, col.names = F)
```

5.Extract data based on list of IDs 
```{bash, eval = F}
plink1.9 --bfile plink_output/var_out_merge --extract merged_rsID.txt --make-bed --out plink_output/merged_common_data
```

Last line of code in PLINK produces an error message - Error: No variants remaining after --extract. I've tried changing separators from tab to spaces and removing column and row names to match the format of the files. PLINK manual suggests that the 'extract' function will accept a text file with a list of variant IDs. Recurring error may be a formatting issue or the commands in PLINK may be in the wrong order or missing a step. 

