---
title: Example of data imputation with missMDA
author: Martha Aquino
date: '2018-04-23'
slug: example-of-data-impuation-with-missmda
categories:
  - Experiments
  - Coding
tags:
  - Data Cleaning
  - Imputation
  - PCA
  - R
---

#Introduction 
This entry will show the results of a single imputation using the missMDA package. The missMDA package imputes quantitative variables using principal component analysis (PCA). 

##Methods and Results 
1. load data 
```{r}
phen.data.age <- read.csv('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/NIES_master_database-age.csv')
phen.data.adults<-phen.data.age[phen.data.age$Age.excel>17,]
quant.variables<- c("R.K.value.H", "R.K.Value.H.Axis", "R.K.value.V", "R.K.value.V.Axis", "L.K.value.H",
                    "L.K.value.H.Axis", "L.K.value.V", "L.K.value.V.Axis", "R.Pachimetry", "L.Pachimetry", "R.Axial.Length",
                    "L.Axial.Length", "AC.Depth.R", "AC.Depth.L", "R.IOP.mmHg", "L.IOP.mmHg", "CDR.RE", "CDR.LE")
quant.data.adults<- phen.data.adults[quant.variables]

```

2. install relevant packages 
```{r}
install.packages('missMDA', dependencies = TRUE, repos = "http://cran.us.r-project.org")
require(missMDA)
require(FactoMineR)
```

3. Estimate the number of dimensions to be used for the reconstruction formula 
```{r}
data(quant.data.adults)
nb = estim_ncpPCA(quant.data.adults)
nb
```

4. Impute the data with the number of dimensions previously calculated 
```{r}
res.comp = imputePCA(quant.data.adults,ncp=nb$ncp)
head(res.comp$completeObs) #view imputed dataset
```

5. Perform PCA on the imputed data set and plot uncertainties
```{r}
res.pca = PCA(res.comp$completeObs)
```

6. Perform multiple imputations 

"res.comp = MIPCA(quant.data.adults, ncp = nb$ncp, nboot = 1000)"

##Discussion 
The single imputation has produced a complete dataset. It is difficult to interpret the individuals factor map in detail because there are a large number of individuals in the study, and each dot represents an individual. If any of the participants had a signifiant amount of data missing, a circle around the dot would be present signifying the uncertainty of the imputed values. The larger the area of the circle, the higher the uncertainty. However, there does not appear to be any that are clearly visible in this plot, which suggests that the imputed values for each individual fits well. It also possible that any circles that are present are not visible because of the dense cluster of dots in the plot. 
The variables factor map shows the uncertainty of the imputed values for each variable.

https://cran.r-project.org/web/packages/missMDA/missMDA.pdf
http://factominer.free.fr/missMDA/PCA.html
http://www.statpower.net/Content/312/R%20Stuff/PCA.html

The results of this PCA imputation does not show the variation for each individual or variable as it was only performed once. Variation, or uncertainty, of the imputed values can be determined by performing multiple imputations. 