---
title: Finding outliers in quantitative data
author: Martha Aquino
date: '2018-04-16'
slug: finding-outliers-in-quantitative-data
categories:
  - Coding
  - Experiments
tags:
  - Data Cleaning
  - R
---

##Introduction 
Finding outliers in a dataset help identify any non-sense data that may not be obvious from the histograms. If they are not non-sense, they may be present due to measurement variability or it may indicate an error. Regardless, it is important to identify outliers as they can significantly affect the results of subsequent statistical analyses and it may be necessary to exclude them from the dataset prior to data imputation. 

##Aim
To identify outliers and non-sense data in the quantitative dataset. 

##Methods and Results 

1. Copy the quantitative dataset so that the original dataset will not be altered 
```{r}
phen.data <- read.csv('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/NIES_master_database.csv')
quant.data <- c("Current.height", "Current.weight", "Logmar.VA.Right", "RVA.with.PH", "Logmar.VA.Left", "LVA.with.PH", 
"R.Sph..pre.dilate.", "R.Cyl..pre.dilate.", "R.Axis..pre.dilate.", "L.Sph..pre.dilate.","L.Cyl..pre.dilate.",
"L.Axis..pre.dilate.", "R.K.value.H", "R.K.Value.H.Axis", "R.K.value.V", "R.K.value.V.Axis", "L.K.value.H",
"L.K.value.H.Axis", "L.K.value.V", "L.K.value.V.Axis", "Lang.score", "R.Pachimetry", "L.Pachimetry", "R.Axial.Length", "L.Axial.Length", "AC.Depth.R", "AC.Depth.L", "R.IOP.mmHg", "L.IOP.mmHg", "CDR.RE", "CDR.LE")
quant.data.subset <- phen.data[quant.data]

quant.data.sub2 <- quant.data.subset
```

2. Try 'extremevalues' package to identify outliers 
```{r}
install.packages('extremevalues', dependencies = TRUE, repos = "http://cran.us.r-project.org" )
require(extremevalues)

outRight<-list()
outLeft<-outRight
for(i in 1:ncol(quant.data.sub2)){
  outRight[[i]]<-getOutliers(quant.data.sub2[,i])$iRight
  outLeft[[i]]<-getOutliers(quant.data.sub2[,i])$iLeft
  quant.data.sub2[outRight[[i]],i]<- NA
  quant.data.sub2[outLeft[[i]],i]<-NA
}

#check output
outLeft
outRight
head(quant.data.sub2)
```

  
3. Try boxplot statistics to identify outliers
```{r}
quant.data.sub3 <- quant.data.subset
outvalue <-list()
outindex <-list()
for (i in 1:ncol(quant.data.sub3)){
  outvalue[[i]] <-boxplot.stats(quant.data.sub3[,i])$out
  outindex[[i]] <-which(quant.data.sub3[,i] == outvalue[[i]])
  quant.data.sub3[outindex[[i]],i] <- NA
}

#check output
outvalue
head(quant.data.sub3)
```

- Each [[number]] represents a quantitative variable from the "quant.data.subset" dataset
- Both methods of identifying an outlier removes the outliers from the dataset and replaces them with "NA".

##Discussion
There are differences in the values identified as outliers using both methods although there are no non-sense data evident. The differences in the outliers identified were due to the the methods employed by the functions. The 'boxplot.stats' determines outliers using the interquartile range (IQR), assuming normal distribution. The 'extremevalues' package estimates a distribution based on Q-Q plots and then determines whether a value would likely fit within the distribution. From the histograms, it was evident that not all of the variables follow a normal distribution, thus using the boxplot statistics may not be the most appropriate method for determining an outlier. Regardless of these results, outliers will not be removed from the dataset. 