---
title: GWAS rerun MAF 0.05
author: Martha Aquino
date: '2018-12-21'
slug: gwas-rerun-maf-0-05
categories:
  - Coding
  - Experiments
tags:
  - Gaston
  - GWAS
  - R
---

```{r}
require(magicfor)
require(magrittr)
require(dplyr)
require(gaston)
require(qqman)
```


```{r}
##load heritable pheno data

her_pheno24918 <- read.csv('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/nies_heritable_pheno240918.csv')
head(her_pheno24918)

nies_covar <- read.csv('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/nies_covar.csv')
head(nies_covar)
```


```{r}
##load genomic data
merged_nies_071118 <- read.bed.matrix("C:/Users/Martha/Documents/Honours/Project/honours.project/Data/merged_nies/merged_nies_geno071118.bed")

```


```{r}
##set GRM etc.
merged_nies_GRM <- GRM(merged_nies_071118)
merged_nies_eiK <- eigen(merged_nies_GRM)
merged_nies_eiK$values[ merged_nies_eiK$values < 0] <- 0
merged_nies_PC <- sweep(merged_nies_eiK$vectors, 2, sqrt(merged_nies_eiK$values), "*") #PCs = 2
```

```{r}
#run GWAS with covariates
for (i in c(2:ncol(her_pheno24918))){
  pheno_colnames <- colnames(her_pheno24918[i])
  her_pheno_gwas <- association.test(merged_nies_071118, her_pheno24918[,i], X = nies_covar, method = "lmm", test = "lrt", K = merged_nies_GRM, eigenK = merged_nies_eiK, p =2)
  her_pheno_gwas <- na.omit(her_pheno_gwas)
  pheno_gwas_filtered <- her_pheno_gwas %>% filter(-log10(p)>1) %>% filter(freqA2 < 0.99)
  qqman::manhattan(x = pheno_gwas_filtered, chr = "chr", bp = "pos", p = "p", snp = "id", genomewideline = -log10(1.84e-7), main = paste(pheno_colnames), ylim = c(0,10))
}
```

As expected, removing the rare and low frequency variants significantly reduced the number of significantly associated variants. 

Interestingly, PCs 1,3, 4, and 8 still show associated variants as well as R Cyl, L ACD, and R ACD. Although these results are yet to be filtered to see whether they are well-supported peaks. In addition, the associated variants previously seen in R KVal V and UVAF were not identified in this round of analysis. 
