---
title: Multiple PCA Imputation of Quantitative Phenotypic Data  - updated
author: Martha Aquino
date: '2018-05-04'
slug: multiple-pca-imputation-of-quantitative-phenotypic-data-updated
categories:
  - Coding
  - Experiments
tags:
  - Data Cleaning
  - Imputation
  - PCA
  - R
---
#Introduction
Multiple imputation with PCA performed on larger data set. 

#Methods and Results 
##1. Load data 
```{r}
phen.data.age <- read.csv('C:/Users/Martha/Documents/Honours/Project/honours.project/Data/NIES_master_database-age.csv')

#include logMAR and base-out prism test values 
phen.data.adults<-phen.data.age[phen.data.age$Age.excel>17,]
quant.variables<- c("Logmar.VA.Right", "RVA.with.PH", "Logmar.VA.Left", "LVA.with.PH", "R.Sph..pre.dilate.", "R.Cyl..pre.dilate.",
                    "R.Axis..pre.dilate.", "L.Sph..pre.dilate.", "L.Cyl..pre.dilate.", "L.Axis..pre.dilate.",
                    "R.K.value.H", "R.K.Value.H.Axis", "R.K.value.V", "R.K.value.V.Axis", "L.K.value.H",
                    "L.K.value.H.Axis", "L.K.value.V", "L.K.value.V.Axis", "R.Pachimetry", "L.Pachimetry", "R.Axial.Length",
                    "L.Axial.Length", "AC.Depth.R", "AC.Depth.L", "R.IOP.mmHg", "L.IOP.mmHg", "CDR.RE", "CDR.LE")
quant.data.adults<- phen.data.adults[quant.variables]
summary(quant.data.adults)
```

##2. Subset relevant data (quantitative only, including logMAR and prism test values, and excluding information from minors)
```{r}
phen.data.adults<-phen.data.age[phen.data.age$Age.excel>17,]
quant.variables<- c("Logmar.VA.Right", "RVA.with.PH", "Logmar.VA.Left", "LVA.with.PH", "R.Sph..pre.dilate.", "R.Cyl..pre.dilate.",
                    "R.Axis..pre.dilate.", "L.Sph..pre.dilate.", "L.Cyl..pre.dilate.", "L.Axis..pre.dilate.",
                    "R.K.value.H", "R.K.Value.H.Axis", "R.K.value.V", "R.K.value.V.Axis", "L.K.value.H",
                    "L.K.value.H.Axis", "L.K.value.V", "L.K.value.V.Axis", "R.Pachimetry", "L.Pachimetry", "R.Axial.Length",
                    "L.Axial.Length", "AC.Depth.R", "AC.Depth.L", "R.IOP.mmHg", "L.IOP.mmHg", "CDR.RE", "CDR.LE")
quant.data.adults<- phen.data.adults[quant.variables]
head(quant.data.adults)

```

##3. View summary statistics of data subset 
```{r}
summary(quant.data.adults)
```

##4. Duplicate data set 
```{r}
ocular_data <- quant.data.adults
```

##5. Convert to double matrix to ensure that all data is 'numeric'
```{r}
ocular_data <- as.matrix(ocular_data)
```

##6.Perform imputation
```{r}
require(missMDA)
require(FactoMineR)
nbdim = estim_ncpPCA(ocular_data, method = 'EM', method.cv="Kfold")
nbdim
res.comp = MIPCA(ocular_data, ncp = nbdim$ncp, nboot = 1000)

res.comp$ncp
```

##7. Plot individuals plot to view uncertainty of predicted values 
```{r}
png('indiv_supp_plot.png', width = 3200, height = 3200, res = 150)
plot.MIPCA(res.comp, choice = "ind.supp") 
dev.off()
```


##8. Plot variable factor plot
```{r}
png('var_factor_plot.png', width = 3200, height = 3200, res = 150)
plot.MIPCA(res.comp, choice = "var") 
dev.off()
```

##Discussion 
From the variable factor plot, it is apparent that the logMAR with pinhole correction values (RVA and LVA with PH) have a large spread, much more scattered than all the other variables. This indicates a relatively large uncertainty for the imputed values for those variables. From the summary statistics, these variables have around 200 missing values, whereas all other variables only have up to 16 missing values. For this reason, we have decided to exclude these variables from subsequent analyses. 

Note: this multiple PCA found five components.